---
title: "HW08-P1 Scraping"
author: "Hamza Mahamud"
format:
  html:
    embed-resources: true
---

## Instructions

**Requirements**

Ex 1

- Have a correct answer to who can scrape that data based on the information in robots.txt

Ex 2

- Write a description of what `httr_attr()` does
- Include a code example to demonstrate what it does to it

Ex 3

- Code should result in 2046 rows and 3 columns of data about cheese
- Code should result in unique information about name, url, and whether or not it has an image
- Code should include Sys.sleep(1) to not overtax website

Ex 4

- Code should result in 10 rows and 6 columns of data about 10 cheeses
- Code should result in unique information about cheese identifier (name or url), milk, country, family, type, flavor
- Code should include Sys.sleep(1) to not overtax website


## Goal

Scrape information from <https://www.cheese.com> to obtain a dataset of characteristics about different cheeses, and gain deeper insight into your coding process.


## Ex 1

Locate and examine the `robots.txt` file for <https://www.cheese.com>. Summarize (in words) what you learn from the cheese.com `robots.txt` file.

The `robots.txt` file at <https://www.cheese.com/robots.txt> applies to all user agents (`User-agent: *`) and contains a single `Disallow: /search/` directive, meaning automated crawlers are asked not to scrape the search functionality. All other pages on the site — including the alphabetical listing at `/alphabetical/` and individual cheese pages — are implicitly allowed. There is no `Crawl-delay` entry, so there is no required pause between requests, though we add one anyway as a courtesy to the server.


## Ex 2

Learn about the `html_attr()` function from `rvest`. Describe how this function works, and give a small code example to illustrate how it works.

`html_attr()` extracts the value of a named HTML attribute from one or more selected nodes. After using `html_elements()` to select a set of nodes, you pass the result to `html_attr("attribute_name")` to pull out whatever value that attribute holds on each node. A common use case is extracting `href` from `<a>` tags to get URLs, or `src` from `<img>` tags to get image paths. If a node does not have the requested attribute, `html_attr()` returns `NA` for that element, which makes it easy to detect missing values.

```{r}
library(rvest)
library(tidyverse)

# Small self-contained example: parse a snippet of HTML and extract href attributes
html_snippet <- read_html('
  <ul>
    <li><a href="https://www.cheese.com/gouda/">Gouda</a></li>
    <li><a href="https://www.cheese.com/brie/">Brie</a></li>
    <li><span>No link here</span></li>
  </ul>
')

# html_attr("href") returns the href value from each <a> node; NA if absent
html_snippet |>
  html_elements("a") |>
  html_attr("href")
```


## Ex 3

Obtain the following information for **all** cheeses in the alphabetical database <https://www.cheese.com/alphabetical/>:

- Cheese name
- URL for the cheese's webpage (e.g., <https://www.cheese.com/gouda/>)
- Whether or not the cheese has a picture (e.g., [gouda](https://www.cheese.com/gouda/) has a picture, but [bianco](https://www.cheese.com/bianco/) does not)

To be kind to the website owners, please add a 1 second pause between page queries. (Note that you can view 100 cheeses at a time.)


```{r}
library(rvest)
library(tidyverse)

# The alphabetical listing paginates 100 cheeses at a time via ?per_page=100&page=N
# We scrape one page at a time and stack the results.

scrape_cheese_page <- function(page_num) {
  Sys.sleep(1)
  page_url <- str_c(
    "https://www.cheese.com/alphabetical/?per_page=100&page=", page_num
  )
  page_html <- tryCatch(read_html(page_url), error = function(e) NULL)
  if (is.null(page_html)) return(NULL)

  # Each cheese is in a <div class="product-item"> card
  cards <- page_html |> html_elements("div.product-item")

  if (length(cards) == 0) return(NULL)

  # Cheese name: text of <h3>
  names <- cards |>
    html_element("h3") |>
    html_text(trim = TRUE)

  # URL: href on the <a> inside <h3>; prepend base URL
  hrefs <- cards |>
    html_element("h3 a") |>
    html_attr("href")
  cheese_urls <- str_c("https://www.cheese.com", hrefs)

  # Has picture: img has class "image-exists" for real photos,
  # "image-missing" for the placeholder SVG
  has_picture <- cards |>
    html_element("img") |>
    html_attr("class") |>
    (\(cls) !is.na(cls) & str_detect(cls, "image-exists"))()

  tibble(name = names, url = cheese_urls, has_picture = has_picture)
}

# Scrape pages until we get an empty page (cheese.com has ~21 pages of 100)
cheese_pages <- map(1:25, scrape_cheese_page)

cheese_df <- cheese_pages |>
  compact() |>       # drop any NULL (empty) pages
  list_rbind() |>
  filter(!is.na(name))  # drop any stray NA rows
```

Print the dimensions & first few rows of your final data frame:

```{r}
dim(cheese_df)
head(cheese_df)
```



## Ex 4

When you go to a particular cheese's page (like [gouda](https://www.cheese.com/gouda/)), you'll see more detailed information about the cheese. For **just 10** of the cheeses in the database, obtain the following detailed information: milk information, country of origin, family, type, and flavour. (Just 10 to avoid overtaxing the website. Continue adding a 1 second pause between page queries.)

```{r}
# Function to scrape detail fields from a single cheese page
scrape_cheese_detail <- function(cheese_url) {
  Sys.sleep(1)
  page_html <- read_html(cheese_url)

  # Helper: extract a labelled fact by its bold-text keyword
  get_fact <- function(keyword) {
    # Look for <p> tags whose text contains the keyword label
    matches <- page_html |>
      html_elements("p") |>
      keep(~ str_detect(html_text(.x), regex(keyword, ignore_case = TRUE)))
    if (length(matches) == 0) return(NA_character_)
    # Strip the label from the full text
    full_text <- html_text(matches[[1]], trim = TRUE)
    str_trim(str_remove(full_text, regex(paste0(".*", keyword, ":?\\s*"), ignore_case = TRUE)))
  }

  tibble(
    url     = cheese_url,
    milk    = get_fact("Made from"),
    country = get_fact("Country of origin"),
    family  = get_fact("Family"),
    type    = get_fact("Type"),
    flavor  = get_fact("Flavou?r")
  )
}

# Take the first 10 cheeses from cheese_df
detail_df <- map(cheese_df$url[1:10], scrape_cheese_detail) |>
  list_rbind()
```

Print the dimensions & first few rows of your final data frame:

```{r}
dim(detail_df)
head(detail_df)
```


## Resources Reflection (required)

List the resources you used to help with this assignment then write 3-5 sentence reflection on which resources were most helpful in finishing this task.

**Resources:**

- `rvest` package documentation: <https://rvest.tidyverse.org/>
- SelectorGadget Chrome extension: <https://chromewebstore.google.com/detail/selectorgadget/mhjhnkcfbdhnjickkkdbjoemdmbfginb>
- cheese.com robots.txt: <https://www.cheese.com/robots.txt>

**Reflection:**

The `rvest` documentation was the most essential resource, especially the `html_elements()` / `html_element()` distinction — using the singular form inside a loop over article containers guarantees aligned vector lengths even when a field is missing on some cards. SelectorGadget made it fast to identify the CSS class for each card and the `<h3>` / `<a>` / `<img>` structure within them, which would have taken much longer by reading raw HTML. Checking `robots.txt` first was a good habit — knowing that only `/search/` is disallowed gave confidence that scraping the alphabetical listing and individual cheese pages was permitted.
